---
title: 'Epidemiology : poster presentation assignment (group 7)'
author: "Boaz Baliejukia"
date: "2024-10-19"
output:
  pdf_document: default
  html_document: default
---
# Loading packages

```{r, message=FALSE}
library(tidyverse)
library(knitr)
```
# Let's upload the dataset

```{r}
sample <- read.csv("Poster Presentation.csv")
head(sample, 15) %>% 
  kable(align = "c")
glimpse(sample)
```

# We are now cleaning data
    
## Let's recode variables
    
```{r}
sample$id <- as.character(sample$id)
sample$sex <- factor(sample$sex, levels = c(0, 1), labels = c("Male", "Female"))
sample$bmi_cat <- factor(sample$bmi_cat, levels = c(1, 2, 3, 4, 5), 
                         labels = c("Underweight", "Normal", "Overweight", 
                                    "Obesity", "Severe obesity"))
sample$bp_baseline <- factor(sample$bp_baseline, levels = c(0, 1), 
                             labels = c("No", "Yes"))
sample$lost <- factor(sample$lost, levels = c(0, 1), 
                             labels = c("No", "Yes"))
sample$bp_followup <- factor(sample$bp_followup, levels = c(0, 1), 
                             labels = c("No", "Yes"))

head(sample, 10) %>% 
  kable(align = "c")

```


## Let's get our population at risk and call it "cohort": adults without LBP at the beginning of the study (bp_baseline = No) 


```{r}
cohort <- sample[sample$bp_baseline == "No", ]
head(cohort, 10) %>% 
  kable(align = "c")

```

### Table 1: Baseline characteristics according to BMI categories

```{r}
cat <- cohort %>%
    group_by(bmi_cat) %>%
    summarise(
        count = n(),
        mean_age = mean(age, na.rm = TRUE),
        sd_age = sd(age, na.rm = TRUE),
        male_percentage = sum(sex == "Male") / n() * 100,
        female_percentage = sum(sex == "Female") / n() * 100
        )
cat %>% 
  kable(align = "c")
```

### Figure 1: Barplot

```{r}

graph1 <- cohort %>%
    drop_na() %>%
    ggplot(aes(bmi_cat, fill = sex)) +
    geom_bar(position = "dodge", alpha = 0.8) +
    theme_classic(base_size = 13) + 
  labs(title = "Participants by BMI groups and gender at baseline", 
       x = "BMI categories", y = "Number of participants")

graph1
```

## We're now splitting our cohort into 2 groups: exposed (obesity) and non-exposed

```{r}
exposed <- cohort[cohort$bmi_cat =="Obesity", ]
non_exposed <- cohort[!cohort$bmi_cat =="Obesity", ]

nrow(exposed)

nrow(non_exposed)

```


### Baseline Demographic characteristics of participants:

```{r}

demo_expo <- exposed %>%
    summarise(
        count = n(),
        mean_age = mean(age, na.rm = TRUE),
        sd_age = sd(age, na.rm = TRUE),
        male_percentage = sum(sex == "Male") / n() * 100,
        female_percentage = sum(sex == "Female") / n() * 100
    )
demo_expo %>% 
  kable(align = "c")

demo_non_expo <- non_exposed %>%
    summarise(
        count = n(),
        mean_age = mean(age, na.rm = TRUE),
        sd_age = sd(age, na.rm = TRUE),
        male_percentage = sum(sex == "Male") / n() * 100,
        female_percentage = sum(sex == "Female") / n() * 100
    )
demo_non_expo %>% 
  kable(align = "c")
```

# Result 

## Attrition bias (lost to follow up):

```{r}
lost_expo <- sum(is.na(exposed$bp_followup))
lost_non_expo <- sum(is.na(non_exposed$bp_followup))

lost_expo
lost_non_expo
```


## Let's create a 2x2 table

Considering :
a = exposed with outcome
b = exposed without outcome
c = non-exposed with outcome
d = non-exposed without outcome


```{r}
a <- length(which(exposed$bp_followup == "Yes"))
b <- length(which(exposed$bp_followup == "No"))
c <- length(which(non_exposed$bp_followup == "Yes"))
d <- length(which(non_exposed$bp_followup == "No"))

a
b
c
d
```
We are now creating a 2x2 matrix :

```{r}
table <- matrix(c(a, b, c, d), nrow = 2, byrow = TRUE)
table <- cbind(table, c(a + b, c + d))
colnames(table) <- c("Low back pain", "Without LBP", "Total")
rownames(table) <- c("Obesity", "Without Obesity")

table %>% 
  kable(align = "c")
write.csv(table, "2x2table.csv")
```

## Let's compute the Risk Ratio

### Risk of outcome in exposed and non-exposed groups
```{r}
lbp_expo <- a/(a+b)

lbp_expo

lbp_non_exp <- c/(c+d)

lbp_non_exp
```

### Risk Ratio
```{r}
RR <- lbp_expo/lbp_non_exp

RR
```

### Figure 2: Barplot

Let's plot a bar graph showing incidence proportion of LBP by exposure groups:

```{r}

cohort$exposure <- factor(
  ifelse(cohort$bmi_cat == "Obesity", "exposed (BMI >= 30 kg/m2)",
  ifelse(cohort$bmi_cat %in% c("Normal", "Overweight", "Underweight"), 
         "non-exposed (BMI < 30 kg/m2)", NA))
)

#cohort
cohort %>%
  drop_na(bp_followup) %>% 
  ggplot(aes(y= exposure, fill = bp_followup)) +
  geom_bar() +
  theme_classic() +
  labs(title = "Incidence of Low Back Pain by exposure groups (Obesity)", 
       x = "Number of participants", y = "Exposure groups")
```

## Table 2: Summary table

```{r}
table1 <- data.frame(
  characteristics = c("Number of participants", 
                               "Mean age",
                               "SD",
                               "Male", "M %", 
                               "Female", "F %", 
                               "Low Back Pain", "LBP %", "LBP M", "LBPM %", "LBP F", "LBPF %"),
  Exposed = c(nrow(exposed), 
                       round(mean(exposed$age), 2), round(sd(exposed$age), 2), 
                       length(which(exposed$sex == "Male")), 
              round(length(which(exposed$sex == "Male")) / nrow(exposed) * 100, 2), 
                       length(which(exposed$sex == "Female")), 
              round(length(which(exposed$sex == "Female")) / nrow(exposed) * 100, 2),
                       a, round(a/nrow(exposed) * 100, 2),
              length(which(exposed$sex == "Male" & exposed$bp_followup == "Yes")), 
              round(length(which(exposed$sex == "Male" & exposed$bp_followup == "Yes")) / a * 100, 2),
              length(which(exposed$sex == "Female" & exposed$bp_followup == "Yes")), 
              round(length(which(exposed$sex == "Female" & exposed$bp_followup == "Yes")) / a * 100, 2)),
  Non_exposed = c(nrow(non_exposed), 
                       round(mean(non_exposed$age), 2), round(sd(non_exposed$age), 2), 
                       length(which(non_exposed$sex == "Male")), 
                  round(length(which(non_exposed$sex == "Male")) / nrow(non_exposed) * 100, 2),
                       length(which(non_exposed$sex == "Female")), 
                  round(length(which(non_exposed$sex == "Female")) / nrow(non_exposed) * 100, 2),
                       c, round(c/nrow(non_exposed) * 100, 2), 
                  length(which(non_exposed$sex == "Male" & non_exposed$bp_followup == "Yes")), 
              round(length(which(non_exposed$sex == "Male" & non_exposed$bp_followup == "Yes")) / c * 100, 2),
              length(which(non_exposed$sex == "Female" & non_exposed$bp_followup == "Yes")), 
              round(length(which(non_exposed$sex == "Female" & non_exposed$bp_followup == "Yes")) / c * 100, 2)))

table1 %>% 
  kable(align = "c")

write.csv(table1, "characteristics.csv", row.names = FALSE)
```

## Table 3: age groups

```{r}

table2 <- data.frame(Age_groups = c("25-35",
                          "36-45",
                          "46-55",
                          "56-65",
                          "66-75"), 
           Number_of_participants = c(length(which(cohort$age < 36)),
                                      length(which(cohort$age > 35 & cohort$age < 46)),
                                      length(which(cohort$age > 45 & cohort$age < 56)),
                                      length(which(cohort$age > 55 & cohort$age < 66)),
                                      length(which(cohort$age > 65))
                                      ), 
           Number_of_cases_LBP = c(length(which(cohort$age < 36 & cohort$bp_followup == "Yes")),
                                      length(which(cohort$age > 35 & cohort$age < 46 & cohort$bp_followup == "Yes")),
                                      length(which(cohort$age > 45 & cohort$age < 56 & cohort$bp_followup == "Yes")),
                                      length(which(cohort$age > 55 & cohort$age < 66 & cohort$bp_followup == "Yes")),
                                      length(which(cohort$age > 65 & cohort$bp_followup == "Yes"))),
           Incidence_proportion = c(round(length(which(cohort$age < 36 & cohort$bp_followup == "Yes")) / 
                                      length(which(cohort$age < 36)) * 100, 2),
                                      round(length(which(cohort$age > 35 & cohort$age < 46 & cohort$bp_followup == "Yes")) /
                                      length(which(cohort$age > 35 & cohort$age < 46)) * 100, 2),
                                      round(length(which(cohort$age > 45 & cohort$age < 56 & cohort$bp_followup == "Yes")) /
                                      length(which(cohort$age > 45 & cohort$age < 56)) * 100, 2),
                                      round(length(which(cohort$age > 55 & cohort$age < 66 & cohort$bp_followup == "Yes")) /
                                      length(which(cohort$age > 55 & cohort$age < 66)) * 100, 2),
                                      round(length(which(cohort$age > 65 & cohort$bp_followup == "Yes")) / 
                                      length(which(cohort$age > 65)) * 100, 2)))
table2 %>% 
  kable(align = "c")

write.csv(table2, "agegroups.csv", row.names = FALSE)           
```










